---
title: "Query Analytics Workloads - Feature Engineering & Model Training"
subtitle: "Atharva Patil · AI/ML Engineer"
author: "Team — UCI #493 Query Analytics Workloads"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

# Setup

Run this chunk first — installs missing packages and creates output directories.

```{r setup, message=FALSE, warning=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org"))

pkgs <- c("readr", "dplyr", "car", "randomForest", "xgboost", "FNN",
          "ggplot2", "Metrics", "tidyr")
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p)
}

library(readr);        library(dplyr);       library(car)
library(randomForest); library(xgboost);     library(FNN)
library(ggplot2);      library(Metrics);     library(tidyr)

set.seed(42)
dir.create("../results", showWarnings = FALSE)
dir.create("../models",  showWarnings = FALSE)
```

---

# Chunk 1 — Load Data

```{r load-data, message=FALSE}
range_log   <- read_csv("../data/processed/clean_range_queries_log.csv")
radius_log  <- read_csv("../data/processed/clean_radius_count_log.csv")

cat("Range Queries rows:", nrow(range_log), "| cols:", ncol(range_log), "\n")
cat("Radius Count rows: ", nrow(radius_log), "| cols:", ncol(radius_log), "\n")
glimpse(range_log)
glimpse(radius_log)
```

---

# Chunk 2 — Feature Engineering

Adds three new features:

- `aspect_ratio` — shape of the rectangle (range queries only)
- `interact_area_dist` — non-linear spatial signal (area × dist_centroid)
- `boundary_flag` — 1 if the query footprint crosses the dataset bounding box

```{r feature-engineering}
# ── Range Queries ──────────────────────────────────────────────────────────────
# Derive bounding box from the data itself (mirrors preprocess.R logic)
x_min_r <- min(range_log$X); x_max_r <- max(range_log$X)
y_min_r <- min(range_log$Y); y_max_r <- max(range_log$Y)

range_fe <- range_log %>%
  mutate(
    aspect_ratio       = X_range / (Y_range + 1e-9),
    log_area           = log1p(area),
    interact_area_dist = log_area * dist_centroid,
    boundary_flag      = as.integer(
      (X - X_range / 2 < x_min_r) | (X + X_range / 2 > x_max_r) |
      (Y - Y_range / 2 < y_min_r) | (Y + Y_range / 2 > y_max_r)
    )
  )

cat("Range boundary rows:", sum(range_fe$boundary_flag), "\n")  # expect ~1 373

# ── Radius Count ───────────────────────────────────────────────────────────────
# All 10 000 radius-count queries cross the bounding box (radius >> box extent
# because coords are pre-normalised to ~[0,1]), so boundary_flag is constant
# and excluded from modeling.
radius_fe <- radius_log %>%
  mutate(
    log_area           = log1p(area),
    interact_area_dist = area * dist_centroid
  )

cat("Radius boundary (all cross):", nrow(radius_fe), "rows\n")
```

---

# Chunk 3 — VIF Analysis

Detects multicollinearity; features with VIF > 10 are dropped from **linear** models (kept for tree-based models).

```{r vif-analysis}
# ── Range Queries: VIF on candidate linear features ───────────────────────────
lm_vif_range <- lm(
  log_Count ~ log_area + X_range + Y_range + dist_centroid +
              aspect_ratio + interact_area_dist + boundary_flag,
  data = range_fe
)
vif_range <- vif(lm_vif_range)
print(round(vif_range, 2))

# Features safe for linear model (VIF ≤ 10)
safe_linear_range <- names(vif_range[vif_range <= 10])
cat("\nSafe linear features (Range):", paste(safe_linear_range, collapse = ", "), "\n")

# ── Radius Count: VIF ─────────────────────────────────────────────────────────
# R and log_area are perfectly collinear (area = pi*R^2); interact_area_dist
# is derived from both. Test a reduced set: log_area + dist_centroid + X_norm + Y_norm
lm_vif_radius <- lm(
  log_Count ~ log_area + dist_centroid + X_norm + Y_norm,
  data = radius_fe
)
vif_radius <- vif(lm_vif_radius)
print(round(vif_radius, 2))

safe_linear_radius <- names(vif_radius[vif_radius <= 10])
cat("\nSafe linear features (Radius):", paste(safe_linear_radius, collapse = ", "), "\n")
```

---

# Chunk 4 — Train / Test Split  (80 / 20, seed = 42)

```{r train-test-split}
set.seed(42)

# ── Range ──────────────────────────────────────────────────────────────────────
n_range       <- nrow(range_fe)
train_idx_r   <- sample(n_range, size = floor(0.8 * n_range))
test_idx_r    <- setdiff(seq_len(n_range), train_idx_r)

range_train   <- range_fe[train_idx_r, ]
range_test    <- range_fe[test_idx_r,  ]

# ── Radius Count ───────────────────────────────────────────────────────────────
n_radius      <- nrow(radius_fe)
train_idx_rad <- sample(n_radius, size = floor(0.8 * n_radius))
test_idx_rad  <- setdiff(seq_len(n_radius), train_idx_rad)

radius_train  <- radius_fe[train_idx_rad, ]
radius_test   <- radius_fe[test_idx_rad,  ]

# Export split indices for reproducibility
write_csv(data.frame(index = train_idx_r,   split = "train"), "../results/split_range.csv")
write_csv(data.frame(index = train_idx_rad, split = "train"), "../results/split_radius.csv")

cat("Range  — train:", nrow(range_train),  "| test:", nrow(range_test),  "\n")
cat("Radius — train:", nrow(radius_train), "| test:", nrow(radius_test), "\n")
```

---

# Chunk 5 — Metric Helper

```{r metrics-helper}
# Returns a named vector: RMSE, MAE, R2, MAPE (all on original Count scale)
calc_metrics <- function(actual_log, pred_log, label = "") {
  actual <- expm1(actual_log)
  pred   <- expm1(pred_log)
  pred   <- pmax(pred, 0)   # floor negatives at 0

  rmse_v <- rmse(actual, pred)
  mae_v  <- mae(actual, pred)
  r2_v   <- 1 - sum((actual - pred)^2) / sum((actual - mean(actual))^2)
  mape_v <- mean(abs((actual - pred) / (actual + 1))) * 100  # +1 avoids /0

  if (nchar(label) > 0) {
    cat(sprintf("  %-8s RMSE=%-10.2f MAE=%-10.2f R2=%.4f MAPE=%.2f%%\n",
                label, rmse_v, mae_v, r2_v, mape_v))
  }
  c(RMSE = rmse_v, MAE = mae_v, R2 = r2_v, MAPE = mape_v)
}
```

---

# Chunk 6 — Linear Regression

## 6a. Range Queries (target: log_Count, secondary: log_SUM)

Linear model uses only VIF-safe features to avoid inflated coefficients.

```{r lm-range}
# Features for linear model: exclude X_range, Y_range (collinear with log_area)
lm_feats_range <- c("log_area", "dist_centroid", "interact_area_dist",
                     "aspect_ratio", "boundary_flag")

f_lm_range <- as.formula(
  paste("log_Count ~", paste(lm_feats_range, collapse = " + "))
)

lm_range <- lm(f_lm_range, data = range_train)
summary(lm_range)

cat("\n── Range LM: Count ──\n")
train_pred_lm_r <- predict(lm_range, range_train)
test_pred_lm_r  <- predict(lm_range, range_test)
lm_range_train  <- calc_metrics(range_train$log_Count, train_pred_lm_r, "Train")
lm_range_test   <- calc_metrics(range_test$log_Count,  test_pred_lm_r,  "Test ")

# Secondary: SUM
f_lm_sum <- as.formula(
  paste("log_SUM ~", paste(lm_feats_range, collapse = " + "))
)
lm_range_sum <- lm(f_lm_sum, data = range_train)
cat("── Range LM: SUM ──\n")
lm_sum_test <- calc_metrics(range_test$log_SUM, predict(lm_range_sum, range_test), "Test ")

# Save metrics
write_csv(
  data.frame(
    dataset = "range", model = "linear",
    split   = c("train", "test"),
    RMSE    = c(lm_range_train["RMSE"], lm_range_test["RMSE"]),
    MAE     = c(lm_range_train["MAE"],  lm_range_test["MAE"]),
    R2      = c(lm_range_train["R2"],   lm_range_test["R2"]),
    MAPE    = c(lm_range_train["MAPE"], lm_range_test["MAPE"])
  ),
  "../results/model_linear.csv"
)
cat("\n[OK] results/model_linear.csv written\n")
```

## 6b. Radius Count

```{r lm-radius}
lm_feats_rad <- c("log_area", "dist_centroid", "X_norm", "Y_norm")

f_lm_rad  <- as.formula(
  paste("log_Count ~", paste(lm_feats_rad, collapse = " + "))
)
lm_radius <- lm(f_lm_rad, data = radius_train)
summary(lm_radius)

cat("\n── Radius LM: Count ──\n")
lm_rad_train <- calc_metrics(radius_train$log_Count, predict(lm_radius, radius_train), "Train")
lm_rad_test  <- calc_metrics(radius_test$log_Count,  predict(lm_radius, radius_test),  "Test ")

write_csv(
  data.frame(
    dataset = "radius", model = "linear",
    split   = c("train", "test"),
    RMSE    = c(lm_rad_train["RMSE"], lm_rad_test["RMSE"]),
    MAE     = c(lm_rad_train["MAE"],  lm_rad_test["MAE"]),
    R2      = c(lm_rad_train["R2"],   lm_rad_test["R2"]),
    MAPE    = c(lm_rad_train["MAPE"], lm_rad_test["MAPE"])
  ),
  "../results/model_linear_radius.csv"
)
```

---

# Chunk 7 — Random Forest (500 trees)

> **Note:** RF on 160 000 rows may take 10–20 minutes. Grab a coffee.

## 7a. Range Queries

```{r rf-range, cache=TRUE}
# Tree models keep all spatial features (VIF collinearity is not a concern)
rf_feats_range <- c("log_area", "X_range", "Y_range", "dist_centroid",
                    "aspect_ratio", "interact_area_dist", "boundary_flag",
                    "X_norm", "Y_norm")

X_train_r <- range_train[, rf_feats_range]
X_test_r  <- range_test[,  rf_feats_range]
y_train_r <- range_train$log_Count
y_test_r  <- range_test$log_Count

set.seed(42)
rf_range <- randomForest(
  x       = X_train_r,
  y       = y_train_r,
  ntree   = 500,
  mtry    = floor(sqrt(length(rf_feats_range))),
  importance = TRUE,
  do.trace   = 100   # prints progress every 100 trees
)

cat("\n── RF Range: Count ──\n")
rf_range_train <- calc_metrics(y_train_r, predict(rf_range, X_train_r), "Train")
rf_range_test  <- calc_metrics(y_test_r,  predict(rf_range, X_test_r),  "Test ")

# Feature importance
imp_range <- importance(rf_range, type = 1)   # %IncMSE
fi_range  <- data.frame(
  feature    = rownames(imp_range),
  importance = imp_range[, 1],
  dataset    = "range",
  model      = "randomForest"
) %>% arrange(desc(importance))

print(fi_range)

write_csv(fi_range, "../results/feature_importance.csv")
cat("\n[OK] results/feature_importance.csv written\n")
```

## 7b. Radius Count

```{r rf-radius, cache=TRUE}
rf_feats_rad <- c("R", "log_area", "dist_centroid",
                  "interact_area_dist", "X_norm", "Y_norm")

set.seed(42)
rf_radius <- randomForest(
  x       = radius_train[, rf_feats_rad],
  y       = radius_train$log_Count,
  ntree   = 500,
  mtry    = floor(sqrt(length(rf_feats_rad))),
  importance = TRUE,
  do.trace   = 100
)

cat("\n── RF Radius: Count ──\n")
rf_rad_train <- calc_metrics(radius_train$log_Count, predict(rf_radius, radius_train[, rf_feats_rad]), "Train")
rf_rad_test  <- calc_metrics(radius_test$log_Count,  predict(rf_radius, radius_test[, rf_feats_rad]),  "Test ")
```

---

# Chunk 8 — XGBoost with 5-Fold CV Grid Search

## 8a. Range Queries — Count (primary)

```{r xgb-range-count, cache=TRUE}
xgb_feats_range <- c("log_area", "X_range", "Y_range", "dist_centroid",
                     "aspect_ratio", "interact_area_dist", "boundary_flag",
                     "X_norm", "Y_norm")

dtrain_r <- xgb.DMatrix(
  data  = as.matrix(range_train[, xgb_feats_range]),
  label = range_train$log_Count
)
dtest_r  <- xgb.DMatrix(
  data  = as.matrix(range_test[,  xgb_feats_range]),
  label = range_test$log_Count
)

# Grid search
grid_r <- expand.grid(
  eta       = c(0.05, 0.1, 0.3),
  max_depth = c(4, 6, 8),
  subsample = c(0.7, 0.9)
)

cv_results_r <- vector("list", nrow(grid_r))

for (i in seq_len(nrow(grid_r))) {
  params <- list(
    booster          = "gbtree",
    objective        = "reg:squarederror",
    eta              = grid_r$eta[i],
    max_depth        = grid_r$max_depth[i],
    subsample        = grid_r$subsample[i],
    colsample_bytree = 0.8,
    seed             = 42
  )
  cv <- xgb.cv(
    params               = params,
    data                 = dtrain_r,
    nrounds              = 500,
    nfold                = 5,
    metrics              = "rmse",
    early_stopping_rounds = 20,
    verbose              = 0
  )
  best_rmse  <- min(cv$evaluation_log$test_rmse_mean)
  best_round <- which.min(cv$evaluation_log$test_rmse_mean)
  cv_results_r[[i]] <- data.frame(grid_r[i, ], best_rmse, best_round)
  cat(sprintf("Grid %2d/%d | eta=%.2f depth=%d sub=%.1f | CV-RMSE=%.4f @ round %d\n",
              i, nrow(grid_r),
              grid_r$eta[i], grid_r$max_depth[i], grid_r$subsample[i],
              best_rmse, best_round))
}

cv_summary_r <- bind_rows(cv_results_r)
best_row_r   <- cv_summary_r[which.min(cv_summary_r$best_rmse), ]
cat("\n── Best params (Range / Count) ──\n"); print(best_row_r)

# Train final model with best params
best_params_r <- list(
  booster          = "gbtree",
  objective        = "reg:squarederror",
  eta              = best_row_r$eta,
  max_depth        = best_row_r$max_depth,
  subsample        = best_row_r$subsample,
  colsample_bytree = 0.8,
  seed             = 42
)
set.seed(42)
xgb_range_count <- xgb.train(
  params  = best_params_r,
  data    = dtrain_r,
  nrounds = best_row_r$best_round,
  verbose = 0
)

cat("\n── XGB Range: Count ──\n")
xgb_rc_train <- calc_metrics(range_train$log_Count, predict(xgb_range_count, dtrain_r), "Train")
xgb_rc_test  <- calc_metrics(range_test$log_Count,  predict(xgb_range_count, dtest_r),  "Test ")
```

## 8b. Range Queries — SUM (secondary)

```{r xgb-range-sum, cache=TRUE}
dtrain_sum <- xgb.DMatrix(
  data  = as.matrix(range_train[, xgb_feats_range]),
  label = range_train$log_SUM
)
dtest_sum  <- xgb.DMatrix(
  data  = as.matrix(range_test[,  xgb_feats_range]),
  label = range_test$log_SUM
)

# Reuse best params found for Count (same feature space)
set.seed(42)
xgb_range_sum <- xgb.train(
  params  = best_params_r,
  data    = dtrain_sum,
  nrounds = best_row_r$best_round,
  verbose = 0
)

cat("\n── XGB Range: SUM ──\n")
xgb_rs_train <- calc_metrics(range_train$log_SUM, predict(xgb_range_sum, dtrain_sum), "Train")
xgb_rs_test  <- calc_metrics(range_test$log_SUM,  predict(xgb_range_sum, dtest_sum),  "Test ")
```

## 8c. Radius Count

```{r xgb-radius-count, cache=TRUE}
xgb_feats_rad <- c("R", "log_area", "dist_centroid",
                   "interact_area_dist", "X_norm", "Y_norm")

dtrain_rad <- xgb.DMatrix(
  data  = as.matrix(radius_train[, xgb_feats_rad]),
  label = radius_train$log_Count
)
dtest_rad  <- xgb.DMatrix(
  data  = as.matrix(radius_test[,  xgb_feats_rad]),
  label = radius_test$log_Count
)

grid_rad <- expand.grid(
  eta       = c(0.05, 0.1, 0.3),
  max_depth = c(4, 6),
  subsample = c(0.7, 0.9)
)

cv_results_rad <- vector("list", nrow(grid_rad))
for (i in seq_len(nrow(grid_rad))) {
  params <- list(
    booster          = "gbtree",
    objective        = "reg:squarederror",
    eta              = grid_rad$eta[i],
    max_depth        = grid_rad$max_depth[i],
    subsample        = grid_rad$subsample[i],
    colsample_bytree = 0.8,
    seed             = 42
  )
  cv <- xgb.cv(
    params                = params,
    data                  = dtrain_rad,
    nrounds               = 300,
    nfold                 = 5,
    metrics               = "rmse",
    early_stopping_rounds = 20,
    verbose               = 0
  )
  best_rmse  <- min(cv$evaluation_log$test_rmse_mean)
  best_round <- which.min(cv$evaluation_log$test_rmse_mean)
  cv_results_rad[[i]] <- data.frame(grid_rad[i, ], best_rmse, best_round)
  cat(sprintf("Grid %2d/%d | eta=%.2f depth=%d sub=%.1f | CV-RMSE=%.4f @ round %d\n",
              i, nrow(grid_rad),
              grid_rad$eta[i], grid_rad$max_depth[i], grid_rad$subsample[i],
              best_rmse, best_round))
}

cv_summary_rad <- bind_rows(cv_results_rad)
best_row_rad   <- cv_summary_rad[which.min(cv_summary_rad$best_rmse), ]
cat("\n── Best params (Radius / Count) ──\n"); print(best_row_rad)

best_params_rad <- list(
  booster          = "gbtree",
  objective        = "reg:squarederror",
  eta              = best_row_rad$eta,
  max_depth        = best_row_rad$max_depth,
  subsample        = best_row_rad$subsample,
  colsample_bytree = 0.8,
  seed             = 42
)
set.seed(42)
xgb_radius_count <- xgb.train(
  params  = best_params_rad,
  data    = dtrain_rad,
  nrounds = best_row_rad$best_round,
  verbose = 0
)

cat("\n── XGB Radius: Count ──\n")
xgb_radc_train <- calc_metrics(radius_train$log_Count, predict(xgb_radius_count, dtrain_rad), "Train")
xgb_radc_test  <- calc_metrics(radius_test$log_Count,  predict(xgb_radius_count, dtest_rad),  "Test ")
```

---

# Chunk 9 — KNN Regressor (K = 5, 10, 20)

Uses `FNN::knn.reg` — fast exact KNN.

## 9a. Range Queries

```{r knn-range}
# Normalise features to [0,1] so Euclidean distance is meaningful
norm_col <- function(x) (x - min(x)) / (max(x) - min(x) + 1e-12)

knn_feats_range <- c("log_area", "dist_centroid", "X_norm", "Y_norm",
                     "aspect_ratio", "interact_area_dist")

X_knn_train_r <- as.matrix(range_train[, knn_feats_range] %>% mutate(across(everything(), norm_col)))
X_knn_test_r  <- as.matrix(range_test[,  knn_feats_range] %>%
                   mutate(across(everything(), ~(. - min(range_train[[cur_column()]])) /
                                               (max(range_train[[cur_column()]]) - min(range_train[[cur_column()]]) + 1e-12))))

knn_k_results_r <- list()
best_knn_rmse_r <- Inf;  best_k_r <- 5

for (k in c(5, 10, 20)) {
  pred <- knn.reg(
    train = X_knn_train_r,
    test  = X_knn_test_r,
    y     = range_train$log_Count,
    k     = k
  )$pred
  m <- calc_metrics(range_test$log_Count, pred, paste0("K=", k))
  knn_k_results_r[[as.character(k)]] <- m
  if (m["RMSE"] < best_knn_rmse_r) { best_knn_rmse_r <- m["RMSE"]; best_k_r <- k }
}
cat("\nBest K for Range KNN:", best_k_r, "\n")
```

## 9b. Radius Count

```{r knn-radius}
knn_feats_rad <- c("log_area", "dist_centroid", "X_norm", "Y_norm", "interact_area_dist")

X_knn_train_rad <- as.matrix(radius_train[, knn_feats_rad] %>% mutate(across(everything(), norm_col)))
X_knn_test_rad  <- as.matrix(radius_test[,  knn_feats_rad] %>%
                    mutate(across(everything(), ~(. - min(radius_train[[cur_column()]])) /
                                                (max(radius_train[[cur_column()]]) - min(radius_train[[cur_column()]]) + 1e-12))))

knn_k_results_rad <- list()
best_knn_rmse_rad <- Inf; best_k_rad <- 5

for (k in c(5, 10, 20)) {
  pred <- knn.reg(
    train = X_knn_train_rad,
    test  = X_knn_test_rad,
    y     = radius_train$log_Count,
    k     = k
  )$pred
  m <- calc_metrics(radius_test$log_Count, pred, paste0("K=", k))
  knn_k_results_rad[[as.character(k)]] <- m
  if (m["RMSE"] < best_knn_rmse_rad) { best_knn_rmse_rad <- m["RMSE"]; best_k_rad <- k }
}
cat("\nBest K for Radius KNN:", best_k_rad, "\n")
```

---

# Chunk 10 — Compile All Results

```{r compile-results}
make_row <- function(dataset, model, split, m) {
  data.frame(dataset, model, split,
             RMSE = m["RMSE"], MAE = m["MAE"], R2 = m["R2"], MAPE = m["MAPE"],
             row.names = NULL)
}

# ── Range / Count ──────────────────────────────────────────────────────────────
# Best KNN test metrics
best_knn_pred_r <- knn.reg(
  train = X_knn_train_r, test = X_knn_test_r,
  y = range_train$log_Count, k = best_k_r
)$pred
knn_rc_test  <- calc_metrics(range_test$log_Count,  best_knn_pred_r, "")

best_knn_pred_train_r <- knn.reg(
  train = X_knn_train_r, test = X_knn_train_r,
  y = range_train$log_Count, k = best_k_r
)$pred
knn_rc_train <- calc_metrics(range_train$log_Count, best_knn_pred_train_r, "")

# ── Radius / Count ─────────────────────────────────────────────────────────────
best_knn_pred_rad <- knn.reg(
  train = X_knn_train_rad, test = X_knn_test_rad,
  y = radius_train$log_Count, k = best_k_rad
)$pred
knn_radc_test <- calc_metrics(radius_test$log_Count,  best_knn_pred_rad, "")

best_knn_pred_train_rad <- knn.reg(
  train = X_knn_train_rad, test = X_knn_train_rad,
  y = radius_train$log_Count, k = best_k_rad
)$pred
knn_radc_train <- calc_metrics(radius_train$log_Count, best_knn_pred_train_rad, "")

# ── Assemble table ─────────────────────────────────────────────────────────────
cv_comparison <- bind_rows(
  make_row("range",  "linear",       "train", lm_range_train),
  make_row("range",  "linear",       "test",  lm_range_test),
  make_row("range",  "randomForest", "train", rf_range_train),
  make_row("range",  "randomForest", "test",  rf_range_test),
  make_row("range",  "xgboost",      "train", xgb_rc_train),
  make_row("range",  "xgboost",      "test",  xgb_rc_test),
  make_row("range",  paste0("knn_k", best_k_r), "train", knn_rc_train),
  make_row("range",  paste0("knn_k", best_k_r), "test",  knn_rc_test),
  make_row("radius", "linear",       "train", lm_rad_train),
  make_row("radius", "linear",       "test",  lm_rad_test),
  make_row("radius", "randomForest", "train", rf_rad_train),
  make_row("radius", "randomForest", "test",  rf_rad_test),
  make_row("radius", "xgboost",      "train", xgb_radc_train),
  make_row("radius", "xgboost",      "test",  xgb_radc_test),
  make_row("radius", paste0("knn_k", best_k_rad), "train", knn_radc_train),
  make_row("radius", paste0("knn_k", best_k_rad), "test",  knn_radc_test)
)

print(cv_comparison, digits = 4)
write_csv(cv_comparison, "../results/cv_comparison.csv")
cat("\n[OK] results/cv_comparison.csv written\n")
```

---

# Chunk 11 — Feature Importance Plot

```{r feature-importance-plot, fig.width=8, fig.height=5}
xgb_fi <- xgb.importance(
  feature_names = xgb_feats_range,
  model         = xgb_range_count
)

xgb_fi_df <- as.data.frame(xgb_fi) %>%
  mutate(Feature = reorder(Feature, Gain))

p_fi <- ggplot(xgb_fi_df, aes(x = Feature, y = Gain)) +
  geom_col(fill = "#2E86AB") +
  coord_flip() +
  labs(
    title    = "XGBoost Feature Importance — Range Queries (Count)",
    subtitle = "Gain = average improvement in loss per split",
    x = NULL, y = "Gain"
  ) +
  theme_minimal(base_size = 13)

print(p_fi)
ggsave("../report/figures/fig07_feature_importance.png", p_fi,
       width = 8, height = 5, dpi = 300)
cat("[OK] report/figures/fig07_feature_importance.png saved\n")

# Append XGBoost importance to feature_importance.csv
xgb_fi_export <- xgb_fi_df %>%
  transmute(feature = as.character(Feature), importance = Gain,
            dataset = "range", model = "xgboost")
fi_combined <- bind_rows(fi_range, xgb_fi_export)
write_csv(fi_combined, "../results/feature_importance.csv")
```

---

# Chunk 12 — Save Model Objects

```{r save-models}
saveRDS(lm_range,         "../models/model_linear_range.rds")
saveRDS(lm_radius,        "../models/model_linear_radius.rds")
saveRDS(rf_range,         "../models/model_rf_range.rds")
saveRDS(rf_radius,        "../models/model_rf_radius.rds")
xgb.save(xgb_range_count, "../models/model_xgb_range_count.model")
xgb.save(xgb_range_sum,   "../models/model_xgb_range_sum.model")
xgb.save(xgb_radius_count,"../models/model_xgb_radius_count.model")

# Save best-K info and training matrices for KNN (KNN needs training data at predict time)
saveRDS(list(k = best_k_r,   X_train = X_knn_train_r,   y_train = range_train$log_Count),
        "../models/model_knn_range.rds")
saveRDS(list(k = best_k_rad, X_train = X_knn_train_rad, y_train = radius_train$log_Count),
        "../models/model_knn_radius.rds")

# Save XGBoost best hyperparameters
xgb_best_params <- bind_rows(
  mutate(best_row_r,   dataset = "range",  target = "Count"),
  mutate(best_row_rad, dataset = "radius", target = "Count")
)
write_csv(xgb_best_params, "../results/xgb_best_params.csv")

cat("[OK] All model objects saved to models/\n")
cat("[OK] XGBoost best params saved to results/xgb_best_params.csv\n")
list.files("../models/")
```

---

# Chunk 13 — Model Comparison Table (for report)

```{r model-table}
# Pretty comparison table filtered to test set only
test_summary <- cv_comparison %>%
  filter(split == "test") %>%
  arrange(dataset, RMSE) %>%
  mutate(across(c(RMSE, MAE, MAPE), ~ round(., 2)),
         R2 = round(R2, 4))

knitr::kable(
  test_summary,
  caption = "Table 1. Test-Set Performance — All Models (Count target, original scale)",
  col.names = c("Dataset", "Model", "Split", "RMSE", "MAE", "R²", "MAPE (%)")
)
```

---

# Summary

```{r session-info}
cat("── Best models by test RMSE ──\n")
best_per_dataset <- test_summary %>%
  group_by(dataset) %>%
  slice_min(RMSE, n = 1)
print(best_per_dataset)

cat("\n── Deliverables written ──\n")
cat("  results/cv_comparison.csv\n")
cat("  results/feature_importance.csv\n")
cat("  results/model_linear.csv\n")
cat("  results/xgb_best_params.csv\n")
cat("  results/split_range.csv  results/split_radius.csv\n")
cat("  models/  (7 model files)\n")
cat("  report/figures/fig07_feature_importance.png\n")

sessionInfo()
```
