---
title: "Query Analytics Workloads — Validation & Diagnostics"
subtitle: "Prasanna Renapurkar · Validation & Report Integrator"
author: "Team — UCI #493 Query Analytics Workloads"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

# Setup

Run this chunk first — installs missing packages and creates output directories.

```{r setup, message=FALSE, warning=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org"))

pkgs <- c("readr", "dplyr", "ggplot2", "xgboost", "cluster",
          "Metrics", "tidyr", "viridis", "gridExtra", "scales")
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p)
}

library(readr);  library(dplyr);   library(ggplot2)
library(xgboost); library(cluster); library(Metrics)
library(tidyr);  library(viridis); library(gridExtra); library(scales)

set.seed(42)
dir.create("../results",        showWarnings = FALSE)
dir.create("../report/figures", showWarnings = FALSE, recursive = TRUE)
```

---

# Chunk 1 — Load Data + Feature Engineering

Mirrors `03_modeling.Rmd` Chunks 1–2 exactly so the validation re-derives the
same feature matrices the trained models saw.

```{r load-fe, message=FALSE}
range_log  <- read_csv("../data/processed/clean_range_queries_log.csv")
radius_log <- read_csv("../data/processed/clean_radius_count_log.csv")

# Range FE
x_min_r <- min(range_log$X); x_max_r <- max(range_log$X)
y_min_r <- min(range_log$Y); y_max_r <- max(range_log$Y)

range_fe <- range_log %>%
  mutate(
    aspect_ratio       = X_range / (Y_range + 1e-9),
    log_area           = log1p(area),
    interact_area_dist = log_area * dist_centroid,
    boundary_flag      = as.integer(
      (X - X_range / 2 < x_min_r) | (X + X_range / 2 > x_max_r) |
      (Y - Y_range / 2 < y_min_r) | (Y + Y_range / 2 > y_max_r)
    )
  )

# Radius FE — boundary_flag constant for radius queries, omitted
radius_fe <- radius_log %>%
  mutate(
    log_area           = log1p(area),
    interact_area_dist = area * dist_centroid
  )

cat("Range :", nrow(range_fe),  "rows\n")
cat("Radius:", nrow(radius_fe), "rows\n")
```

---

# Chunk 2 — Reproduce Train / Test Split (seed = 42)

Saved model files in `models/` are Git-LFS pointer stubs in this checkout, so
we re-create the 80/20 split with the same seed used in `03_modeling.Rmd`
and refit XGBoost from `xgb_best_params.csv`. The split is identical to the
one in `results/split_range.csv` / `results/split_radius.csv`.

```{r split}
set.seed(42)
n_range       <- nrow(range_fe)
train_idx_r   <- sample(n_range, size = floor(0.8 * n_range))
test_idx_r    <- setdiff(seq_len(n_range), train_idx_r)
range_train   <- range_fe[train_idx_r, ]
range_test    <- range_fe[test_idx_r,  ]

n_radius      <- nrow(radius_fe)
train_idx_rad <- sample(n_radius, size = floor(0.8 * n_radius))
test_idx_rad  <- setdiff(seq_len(n_radius), train_idx_rad)
radius_train  <- radius_fe[train_idx_rad, ]
radius_test   <- radius_fe[test_idx_rad,  ]

cat(sprintf("Range  : train=%d  test=%d\n", nrow(range_train),  nrow(range_test)))
cat(sprintf("Radius : train=%d  test=%d\n", nrow(radius_train), nrow(radius_test)))
```

---

# Chunk 3 — Refit XGBoost (Range + Radius)

```{r refit-xgb, cache=TRUE}
best_params <- read_csv("../results/xgb_best_params.csv", show_col_types = FALSE)
br_range  <- best_params %>% filter(dataset == "range")  %>% slice(1)
br_radius <- best_params %>% filter(dataset == "radius") %>% slice(1)

xgb_feats_range <- c("log_area", "X_range", "Y_range", "dist_centroid",
                     "aspect_ratio", "interact_area_dist", "boundary_flag",
                     "X_norm", "Y_norm")
xgb_feats_rad   <- c("R", "log_area", "dist_centroid",
                     "interact_area_dist", "X_norm", "Y_norm")

dtrain_r   <- xgb.DMatrix(as.matrix(range_train [, xgb_feats_range]), label = range_train$log_Count)
dtest_r    <- xgb.DMatrix(as.matrix(range_test  [, xgb_feats_range]), label = range_test $log_Count)
dtrain_rad <- xgb.DMatrix(as.matrix(radius_train[, xgb_feats_rad]),   label = radius_train$log_Count)
dtest_rad  <- xgb.DMatrix(as.matrix(radius_test [, xgb_feats_rad]),   label = radius_test $log_Count)

set.seed(42)
xgb_range <- xgb.train(
  params  = list(objective = "reg:squarederror",
                 eta = br_range$eta, max_depth = br_range$max_depth,
                 subsample = br_range$subsample, colsample_bytree = 0.8, seed = 42),
  data    = dtrain_r,
  nrounds = br_range$best_round,
  verbose = 0
)

set.seed(42)
xgb_radius <- xgb.train(
  params  = list(objective = "reg:squarederror",
                 eta = br_radius$eta, max_depth = br_radius$max_depth,
                 subsample = br_radius$subsample, colsample_bytree = 0.8, seed = 42),
  data    = dtrain_rad,
  nrounds = br_radius$best_round,
  verbose = 0
)

# Predictions and residuals on test sets (log scale + original scale)
range_test  <- range_test  %>% mutate(
  pred_log  = predict(xgb_range,  dtest_r),
  actual    = expm1(log_Count),
  pred      = pmax(0, expm1(pred_log)),
  resid_log = log_Count - pred_log,
  resid     = actual - pred
)
radius_test <- radius_test %>% mutate(
  pred_log  = predict(xgb_radius, dtest_rad),
  actual    = expm1(log_Count),
  pred      = pmax(0, expm1(pred_log)),
  resid_log = log_Count - pred_log,
  resid     = actual - pred
)

cat(sprintf("Range  XGB test RMSE = %10.2f  R^2 = %.4f\n",
            rmse(range_test$actual,  range_test$pred),
            1 - sum((range_test$actual  - range_test$pred )^2) /
                sum((range_test$actual  - mean(range_test$actual ))^2)))
cat(sprintf("Radius XGB test RMSE = %10.2f  R^2 = %.4f\n",
            rmse(radius_test$actual, radius_test$pred),
            1 - sum((radius_test$actual - radius_test$pred)^2) /
                sum((radius_test$actual - mean(radius_test$actual))^2)))
```

---

# Chunk 4 — Final Benchmark (`results/final_benchmark.csv`)

Reshapes `results/cv_comparison.csv` into a single train+test side-by-side table
(deliverable named in the sprint plan).

```{r final-benchmark, message=FALSE}
cv <- read_csv("../results/cv_comparison.csv", show_col_types = FALSE)

final_bench <- cv %>%
  pivot_wider(names_from = split,
              values_from = c(RMSE, MAE, R2, MAPE),
              names_sep = "_") %>%
  arrange(dataset, RMSE_test)

write_csv(final_bench, "../results/final_benchmark.csv")
cat("[OK] results/final_benchmark.csv written\n")

knitr::kable(final_bench, digits = 4,
             caption = "Final benchmark — train and test metrics, all models")
```

---

# Chunk 5 — M4: Residual Diagnostics for XGBoost

Four-panel diagnostic figure:
1. Range — fitted vs residual (log scale) with LOESS overlay → heteroscedasticity check
2. Radius — fitted vs residual (log scale)
3. Range — predicted vs actual (original Count scale)
4. Radius — predicted vs actual

Test sets are sub-sampled for plotting; metrics use the full test set.

```{r residuals-xgb, fig.width=12, fig.height=9}
set.seed(42)
samp_r  <- sample_n(range_test,  size = min(5000, nrow(range_test)))
samp_rd <- sample_n(radius_test, size = min(2000, nrow(radius_test)))

p1 <- ggplot(samp_r, aes(pred_log, resid_log)) +
  geom_point(alpha = 0.3, size = 0.6, color = "#2E86AB") +
  geom_hline(yintercept = 0, color = "red", linewidth = 0.5) +
  geom_smooth(method = "loess", color = "darkorange", se = FALSE, linewidth = 0.7) +
  labs(title = "Range Queries — XGBoost Residuals (log scale)",
       x = "Predicted log_Count", y = "Residual") +
  theme_minimal(base_size = 12)

p2 <- ggplot(samp_rd, aes(pred_log, resid_log)) +
  geom_point(alpha = 0.4, size = 0.7, color = "#E63946") +
  geom_hline(yintercept = 0, color = "red", linewidth = 0.5) +
  geom_smooth(method = "loess", color = "darkorange", se = FALSE, linewidth = 0.7) +
  labs(title = "Radius Count — XGBoost Residuals (log scale)",
       x = "Predicted log_Count", y = "Residual") +
  theme_minimal(base_size = 12)

p3 <- ggplot(samp_r, aes(actual, pred)) +
  geom_point(alpha = 0.3, size = 0.6, color = "#2E86AB") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(title = "Range Queries — Predicted vs Actual",
       x = "Actual Count", y = "Predicted Count") +
  theme_minimal(base_size = 12)

p4 <- ggplot(samp_rd, aes(actual, pred)) +
  geom_point(alpha = 0.4, size = 0.7, color = "#E63946") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Radius Count — Predicted vs Actual",
       x = "Actual Count", y = "Predicted Count") +
  theme_minimal(base_size = 12)

fig08 <- arrangeGrob(p1, p2, p3, p4, ncol = 2)
ggsave("../report/figures/fig08_residuals_xgb.png", fig08,
       width = 12, height = 9, dpi = 300)
grid::grid.draw(fig08)
cat("[OK] report/figures/fig08_residuals_xgb.png saved\n")
```

---

# Chunk 6 — D3 (part 1): Bias by `log_area` Decile (Range)

XGBoost test-set RMSE is computed within each `log_area` decile to surface where
the model under-performs.

```{r bias-quantile, fig.width=8, fig.height=5}
range_test_b <- range_test %>%
  mutate(area_decile = ntile(log_area, 10))

bias_decile <- range_test_b %>%
  group_by(area_decile) %>%
  summarise(
    n           = n(),
    min_area    = min(area),
    max_area    = max(area),
    mean_actual = mean(actual),
    rmse        = rmse(actual, pred),
    mae         = mae(actual, pred),
    mape        = mean(abs((actual - pred) / (actual + 1))) * 100,
    .groups     = "drop"
  ) %>%
  arrange(area_decile)

write_csv(bias_decile, "../results/bias_by_decile.csv")
knitr::kable(bias_decile, digits = 2,
             caption = "XGBoost test-set error by log_area decile (Range)")

p_bias <- ggplot(bias_decile, aes(factor(area_decile), rmse)) +
  geom_col(fill = "#2E86AB") +
  geom_text(aes(label = comma(round(rmse))), vjust = -0.4, size = 3) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(title    = "Range Queries — XGBoost Test RMSE by Area Decile",
       subtitle = "Decile 1 = smallest area, Decile 10 = largest",
       x = "log_area decile", y = "RMSE (Count scale)") +
  theme_minimal(base_size = 13)

ggsave("../report/figures/fig09_bias_by_quantile.png", p_bias,
       width = 8, height = 5, dpi = 300)
print(p_bias)
cat("[OK] report/figures/fig09_bias_by_quantile.png saved\n")
```

---

# Chunk 7 — D1: K-Means K ∈ {3, 4, 5} with Silhouette Selection

K-Means is fit on (X_norm, Y_norm, log_area_norm) for the Range Queries
sub-dataset (largest sample, 200k rows). Silhouette score is computed on a
5,000-row sample because `silhouette()` requires an n × n distance matrix
(O(n²) memory).

```{r kmeans, cache=TRUE}
norm_min_max <- function(x) (x - min(x)) / (max(x) - min(x) + 1e-12)

km_data <- range_fe %>%
  transmute(X_norm,
            Y_norm,
            log_area_norm = norm_min_max(log_area)) %>%
  as.matrix()

set.seed(42)
sil_idx  <- sample(nrow(km_data), 5000)
sil_data <- km_data[sil_idx, ]
sil_dist <- dist(sil_data)

km_results <- list()
sil_scores <- numeric(3)
ks         <- c(3, 4, 5)

for (i in seq_along(ks)) {
  k <- ks[i]
  set.seed(42)
  km_full <- kmeans(km_data, centers = k, nstart = 10, iter.max = 50)
  km_results[[as.character(k)]] <- km_full

  set.seed(42)
  km_sub <- kmeans(sil_data, centers = k, nstart = 10, iter.max = 50)
  sil    <- silhouette(km_sub$cluster, sil_dist)
  sil_scores[i] <- mean(sil[, 3])
  cat(sprintf("K = %d  silhouette = %.4f  WSS = %.2f\n",
              k, sil_scores[i], km_full$tot.withinss))
}

best_k  <- ks[which.max(sil_scores)]
km_best <- km_results[[as.character(best_k)]]
cat("\nBest K (silhouette):", best_k, "\n")

range_fe$cluster <- factor(km_best$cluster)
```

---

# Chunk 8 — D2: 2D Cluster Map

```{r cluster-map, fig.width=8, fig.height=6}
set.seed(42)
samp_clu <- range_fe %>% sample_n(min(15000, n()))

p_clu <- ggplot(samp_clu, aes(X_norm, Y_norm, color = cluster)) +
  geom_point(alpha = 0.5, size = 0.4) +
  scale_color_viridis_d(option = "plasma") +
  coord_equal() +
  labs(title    = paste0("Range Queries — K-Means Clusters (K = ", best_k, ")"),
       subtitle = "Features: X_norm, Y_norm, normalised log_area",
       x = "X (normalised)", y = "Y (normalised)") +
  theme_minimal(base_size = 13) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))

ggsave("../report/figures/fig10_clusters.png", p_clu,
       width = 8, height = 6, dpi = 300)
print(p_clu)
cat("[OK] report/figures/fig10_clusters.png saved\n")
```

---

# Chunk 9 — D3 (part 2): Bias by Spatial Cluster

Joins each test row to its K-Means cluster (via `test_idx_r`) and reports
per-cluster XGBoost RMSE — surfaces which Chicago zones the model handles
worst.

```{r bias-cluster}
range_test_c <- range_test %>%
  mutate(cluster = factor(km_best$cluster[test_idx_r]))

bias_cluster <- range_test_c %>%
  group_by(cluster) %>%
  summarise(
    n              = n(),
    mean_actual    = mean(actual),
    rmse           = rmse(actual, pred),
    mae            = mae(actual, pred),
    mape           = mean(abs((actual - pred) / (actual + 1))) * 100,
    pct_boundary   = mean(boundary_flag) * 100,
    .groups        = "drop"
  ) %>%
  arrange(desc(rmse))

write_csv(bias_cluster, "../results/bias_by_cluster.csv")
knitr::kable(bias_cluster, digits = 2,
             caption = "XGBoost test-set error by K-Means cluster (Range)")
```

---

# Summary

```{r summary}
cat("── Validation deliverables written ──\n")
cat("  results/final_benchmark.csv\n")
cat("  results/bias_by_decile.csv\n")
cat("  results/bias_by_cluster.csv\n")
cat("  report/figures/fig08_residuals_xgb.png\n")
cat("  report/figures/fig09_bias_by_quantile.png\n")
cat("  report/figures/fig10_clusters.png\n\n")

cat(sprintf("Best K (silhouette): %d (silhouette = %.4f)\n",
            best_k, max(sil_scores)))
cat(sprintf("Range  XGB test RMSE: %.2f\n", rmse(range_test$actual,  range_test$pred)))
cat(sprintf("Radius XGB test RMSE: %.2f\n", rmse(radius_test$actual, radius_test$pred)))

sessionInfo()
```
